---
title: "Chapter 9: Box and Whisker"
date: last-modified
date-fomrat: "MMM D"
author: "Dylan M. Ayres, Jacob A. Serrano, Dr. Amy Wiles"
format:
  html:
    theme: slate
    toc: true
    toc-depth: 5
    toc-expand: true
    number-sections: false
---
# Chapter Introduction

In this chapter, we will be discussing how to create a box and whisker plot for the representation of your bioinformatic data. We will also discuss ways that you can adjust the box and whisker plots to show more nuanced aspects of the expression of a gene. First, it is important to introduce the concept of box and whisker plots.

A box and whisker plot is used to show the graphical representation of variation in data. Included in these plots are the minimum value, the maximum value, second quartile, third quartile, and the median value. The visual representation of these can be shown through points on the box and whisker plot or through the lines shown on a box and whisker plot.

## Installing R and Library Loading

Before we can begin to process data, we need to have R and RStudio installed on our computer. The next step is to load your data into the RStudio. To do this, please follow the guide provided in [**File Download Instructions**](File_download.qmd). Afterwards, we need to install some packages. The main package is `tidyverse` which allows for ease of writing and editing our code.\

After completing this step, you should have all the tools necessary to create a simple box and whisker plot with your data. The next step is to learn how to call the function to create a box and whisker plot. Afterwards, we will discuss what functionality `base R` has to customize the box and whisker plot.

```{r Tidyverse, include=FALSE, echo=FALSE, message=FALSE, results='hide', warning=FALSE, error=FALSE}
library(tidyverse)
library(ggplot2)
library(kableExtra)
library(ggrepel)
library(here)
```

# Base R - Box and Whisker

In this section, we will view the simplistic box plot that comes with base R. Afterwards, there will be a note about some small adjustment functions for the customization of the base R box plot code. Following this section will be the discussion of box plots with `ggplot2`, which allows for greater levels of customization.

## Simple Boxplot

Before we get into the customization options for the box and whisker plot, we need to see what a simple box plot code looks like through R.\

To create a simple box and whisker plot in R, you will use the `boxplot` command. To use the base R `boxplot` code, you will need to have a list of your data. For explanation purposes, I will create one below:

```{r Base R - Data Set}
df <- c(10, 12, 14, 16, 20, 25, 30, 41, 8, 15, 31)

df
```

Now that the data is created and assigned to the `df` object, I will use the `boxplot` function on the data. Remember that this is the most basic form of a box plot, and that customization will be in a later step.

```{r Base R - Simple Boxplot}
boxplot(df)
```

\
Now that we have covered how to make a simple box plot, we will discuss how to add customization to the box plots in base R.

## Customization in Base R

In the base `boxplot` code, you can use the following functions to customize its appearance:

-   Notching:
    -   A representation of your confidence intervals for your data
    -   Uses `notch = TRUE`
-   Coloring:
    -   A way to differentiate between two different box plots
    -   Uses `col = c("red", "blue")` (can change to any color)
-   Labels:
    -   A way to indicate the meaning behind your axes scale
    -   Uses `main = "_"`, `xlab = "_"`, and `ylab = "_"`
-   Horizontal:
    -   A way to turn your box plots from vertical to horizontal
    -   Uses `horizontal = TRUE`\

# `ggplot2` {#ggplot2}

Now that we have covered some basic box plots using the base R functions, we can cover more complex box plots using `ggplot2`. The first step is to install the package.\

## Simple `ggplot2` Box Plot {#ggplot2-box-plot}

### Data Creation
```{r date conversion, include=FALSE, eval=FALSE}
# 1. Get the path of the file currently open in RStudio
current_path <- rstudioapi::getActiveDocumentContext()$path

# 2. Extract the directory and base filename
dir_name <- dirname(current_path)
base_name <- "Box and Whisker" # The static part of your name

# 3. Create the new name with today's date (e.g., "Jan_22")
new_name <- paste0(base_name, ".qmd")
new_path <- file.path(dir_name, new_name)

# 4. Rename the file on the disk
if (current_path != new_path) {
  file.rename(current_path, new_path)
  message("File renamed to: ", new_name)
}
```


Now we can move on to creating simple box plots using the `ggplot2` function. Using this function, you will need to have a data frame or matrix format for your data. Below, I will create a subset of data for the purposes of explaining how `ggplot2` uses the data. Here, I create a `Verification Data` object from the `Verification Data.xslx` file. Afterwards, I remove columns with replicates, the zero hour, and the verification position per hour. Then, I remove the extra column numbering the data rows. Finally, I create a smaller subset of the data with the first ten rows of data and the eight columns, consisting of verification position, hour, and the six conditions. Finally, I remove the prefix "norm" and the suffix "avg" from the condition columns.

```{r ggplot2 Data, include=FALSE, echo=FALSE}
Verification_Data <- readxl::read_excel("C:/Users/dylan/OneDrive/Documents/R development/Box_Whisker/B&W Data/Verification Data.xlsx")
df_box <- Verification_Data[-c(1:288),]
delete_1 <- "_rep1$"
delete_2 <- "_rep2$"
delete_3 <- "_hr$"

df_box <- df_box %>% 
  select(-matches(delete_1), -matches(delete_2), -matches(delete_3))
  
df_box <- df_box[,-c(1)]

df_box <- df_box[1:10, 1:8]

df_box <- df_box %>%
  rename_with(~ str_remove_all(., "^norm_"), everything()) %>%
  rename_with(~ str_remove_all(., "_avg$"), everything())

df_box_1 <- df_box %>% 
  head(8)
```

After creating that data set, you will be able to view what the data set looks like as an object. For the sake of brevity, I limited the data above to eight rows.

```{r Table 1, echo=FALSE, results='markup'}
kbl(df_box_1, format = "simple", caption = "Table of Data")
```

Now that the table has been made, we can use the data to create a box plot. The first step is to use the `ggplot` function to initiate a plot and use the `geom_boxplot` to create a box plot. To do this, I will use the function, `pivot_longer`. After pivoting the data frame, I will show you the data.

```{r Pivoted Data}
df_bw <- df_box %>% 
  pivot_longer(cols = !c(1:2),
               names_to = "conditions",
               values_to = "abs")

df_bw_1 <- df_bw %>% 
  head(8)
```

Once again, for the sake of brevity, I have limited this table to the first 8 rows. Though there will be more data when you run the code than shown here.

```{r Truncated Pivoted Data, echo=FALSE, results='markup'}
kbl(df_bw_1, format = "simple", caption = "Truncated Table of Data")
```

Now that we have successfully pivoted the data frame, we can apply it to `ggplot` to create our box plot. To do this, specify what data you want to use after the `ggplot` function, and specify the aesthetic. In this case, I wanted the conditions to be on the x-axis and the absorbance values to be on the y-axis. This is easily interchangeable. Afterwards, add a `+` and write the `geom_boxplot()` line. Then run your code.

```{r Simple `ggplot`}
ggplot(df_bw, aes(x = conditions, y = abs))+
  geom_boxplot()
```

Here, you will see that there is an outlier shown above the "SD30" condition. In a later step, we will explain how you can find which peice of data this outlier is and label it. Though for now, all data is represented by the shown box and whisker plot and its singular outlier.

## Customizing `ggplot2` {#custom-ggplot2}

Using `ggplot`, there are many ways that you can customize your box plot graph. In the following sections, we will discuss many ways that you can change your graphs and what they mean. The first on our list is changing the colors of the box plots.

### Basic Customization

#### Creating an Order

Sometimes, when you upload data to `ggplot`, your order of your data will be moved around. To remedy this, we need to create a custom order for our data frame. To do so, follow the example shown below:

```{r Order Creation}
desired_order <- c("SD30", "SD37", "LN30", "LN37", "LS30", "LS37")

df_bw$conditions <- factor(df_bw$conditions, levels = desired_order)
```

#### `geom_point` {#geompoint}

The first major customization, besides creating an order, that is allowed in `ggplot2` is the ability to create points on your graph. This can be done by using the `geom_point` argument. An example is provided below:

```{r geom_point}
df_bw %>% 
  ggplot(
    aes(x = conditions, y = abs)
  )+
  geom_boxplot()+
  geom_point()
```

After creating the basic points, we can customize this even more! In `ggplot2`, different shape types are described to a certain number. By changing the number, your points in your graph will change. Below is the list for numbers 0-15 and I will use shape 15 in a `ggplot` code for a boxplot.

```         
0. Box
1. Circle
2. Triangle
3. Cross
4. "X"
5. Diamond
6. Triangle (facing downward)
7. Box with "X"
8. Asterisk
9. (Rotated) Box with Cross
10. Circle with Cross
11. Star of David
12. Box with Cross
13. Circle with "X"
14. Box with Traingle
15. Solid Box
```

```{r geom_point 15}
df_bw %>% 
  ggplot(aes(x=conditions, y=abs))+
  geom_boxplot()+
  geom_point(shape = 15)
```

You might notice, though, that you can't tell how many points might be overlapping, this is also the case for the outlier. You are not sure how many outliers are represented. So, first we will cover how to make our outlier stand out, and then how to `jitter` our points on the graph so that we can represent all data points on the graph.

#### Outliers

The same shapes shown in geom_point are used in in `geom_jitter` and the outliers argument of `geom_boxplot`. In the case of the outlier argument, there is more customization options that need to be addressed. For the sake of brevity, I will make a list below of the customization options:\

-   outlier.color/colour: changes the color of the outliers in the box plot
-   outlier.shape: changes the shape of outliers to a specified shape
    -   `outlier.shape = NA`: removes outliers from the graph
-   outlier.size: changes the size of outliers
-   outlier.stroke: changes the thickness and border of outliers
-   outlier.alpha: changes the opacity of outliers

The next step is to create a box and whisker plot with the outlier customized. In this case, I customized the outlier to be a solid red box with a thickness of 1. You should notice, though, that I did not include geom_point in the example below. This is due to the application of the `geom_boxplot` code. The `geom_boxplot` creates an outlier by default, adding lines like `geom_point`, and `geom_jitter` (our next section) we create that point for a second time. In the geom_jitter section, I will explain the fix to this residual artifact.

```{r Outliers}
df_bw %>% 
  ggplot(
    aes(x=conditions,y=abs)
  )+
  geom_boxplot(outlier.shape = 15, outlier.stroke = 1, outlier.colour = "red")
```

##### Outlier Labeling

Another way to customize your outliers in your box plots is to create a custom label for your outlier. In this case, we will be using a new data file, known as `df_test` with more than one outlier in the total data set.


```{r Secondary Outlier Creation, include=FALSE, echo=FALSE}
df_test <- read.csv("dataframe_bw.csv") 
```

The first step is to create a flag for our outliers. To do this, we group the data by our `conditions`, then calculate the Q1 and Q3 levels of absorbance. Afterwards, we calculate the IQR and create the fenced. Then, we check the absorbance of all data to see if there is outlier above or below the fences. Finally, we ungroup the data so it can be used in a later step.

```{r Outlier Identification}
data_with_outlier_flag <- df_test %>%
  group_by(conditions) %>%
  mutate(
    Q1 = quantile(abs, 0.25),
    Q3 = quantile(abs, 0.75),
    IQR = Q3 - Q1,
    lower_fence = Q1 - 1.5 * IQR,
    upper_fence = Q3 + 1.5 * IQR,
    is_outlier = abs < lower_fence | abs > upper_fence
  ) %>%
  ungroup()
```

Now, we need to identify the name of the outlier. In this case, we will use the data frame we made previously and filter to only have the data confirmed as an outlier. Afterwards, we create an outlier label of the verification position of the outlier.

```{r Outlier Naming}
outlier_points_data <- data_with_outlier_flag %>%
  filter(is_outlier) %>%
  mutate(outlier_label = paste0(ver_pos))
```

To put it all together, we will use the same `geom_boxplot` function we have used previously, and now add `geom_text` into the mix. When using `geom_text` we want to specify we are using our new data frame and copy the aesthetics of the boxplots. Then, we add the outlier label to the point. We color it black, just like the point in the boxplot function, and give it a vertical adjustment of negative 1. In this case, `vjust` needs to be negative, as the positive direction is towards 0 and we want the label to be above the point. Finally, we can specify the size of the label.

```{r Outliers labeled}
  ggplot(df_test, 
    aes(x=conditions, y=abs))+
  geom_boxplot(outlier.shape = 15, outlier.color = "black", outlier.stroke = 2)+
  geom_text(data = outlier_points_data,
            aes(x = conditions, y = abs, label = outlier_label),
            color = "black",
            vjust = -1,
            size = 3)

```

#### `geom_jitter` {#geomjitter}

Now that we have covered how to label our outlier on our graph, we will move into jittered points on our graph. To display your data on the box and whisker plots, you will need to use the `geom_jitter` argument. However, when you are using this line, be careful about artifacts. An example of how to use `jitter` is shown below, where I use a distance of 0.3 between the points:

[Jacob or Dylan]{style="color: orange;"}: Explain what jitter does more

```{r Jitter}
df_bw %>% 
  ggplot(
    aes(x = conditions, y = abs)
  )+
  geom_boxplot()+
 geom_jitter(width = 0.3)
```

As you can see, the outlier is represented twice. To check if the outlier is actually written twice in our data set, we can use the `outlier.shape = NA` function we discussed in the previous section. As shown in the example below:

```{r jittered without outlier}
df_bw %>% 
  ggplot(
    aes(x = conditions, y = abs)
  )+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width = 0.2)

```

As you can see, the outlier is only represented once. This means that the one outlier was an artifact from previous code. Be careful about this artifact in future graphs. Now that we have covered some basic customization, we can examine how to change the aesthetics of our graph.

```{r Jitter outlier free data}
df_bw_outlier <- df_bw %>%
  group_by(conditions) %>%
  mutate(
    Q1 = quantile(abs, 0.25),
    Q3 = quantile(abs, 0.75),
    IQR = Q3 - Q1,
    is_outlier = ifelse(abs < (Q1 - 1.5 * IQR) | abs > (Q3 + 1.5 * IQR), TRUE, FALSE)
  ) %>%
  mutate(
    label_text = ifelse(is_outlier, as.character(ver_pos), NA)
  ) %>%
  ungroup()

df_bw_no_outlier <- df_bw_outlier %>% 
  filter(is_outlier == FALSE)

df_bw_outlier_only <- df_bw_outlier %>% 
  filter(is_outlier == TRUE)


```

```{r Jitter with colored & labeled outlier}
df_bw %>% 
  ggplot(aes(x=conditions, y=abs))+
  geom_boxplot(outlier.shape = 15, outlier.color = "red", outlier.stroke =1)+
  geom_jitter(data = df_bw_no_outlier, width = 0.3)+
  geom_text(data = df_bw_outlier_only,
            aes(x = conditions, y = abs, label = label_text),
            color = "black",
            vjust = -0.5,
            size = 3)
```

### `aes` Customization {#aes-custom}

#### Coloring Your Box Plot

To change the colors of your box plots, you will need to use the `color` argument like in the base R. Thought this time, the argument will be placed in the `aes()` line. When doing so, the outline of the box plots will be colored in different shades. An example is shown below:

```{r Color ggplot}
df_bw %>% 
  ggplot(
    aes(x = conditions, y = abs, color = conditions))+
    geom_boxplot()
```

If your want to change the interior color of the box plots, you can do that with the `fill` argument instead of the `color` argument. An example is shown below:

```{r Filled Box Plots}
df_bw %>% 
  ggplot(
    aes(x = conditions, y = abs, fill = conditions)
  )+
  geom_boxplot()
```

With these techniques for coloring in mind, we should also examine how we can demonstrate multiple between multiple conditions. Below is an example of differentiating between temperature and conditions:
##### Multi Filtered
```{r multi filtered data}
df_bw_sep <- df_bw %>% 
  group_by(conditions) %>%
  mutate(
    Condition = str_extract(conditions, "[A-Za-z]+"), 
    Temp = as.numeric(str_extract(conditions, "[0-9]+"))) %>% 
  mutate(
    Condition = factor(Condition, levels = c("SD", "LN", "LS")),
    Temp = factor(Temp, levels = c("30", "37"))
  )

```

```{r multi filtered objects}
common_legend_title <- "Experimental Conditions"

temp_fill_colors <- c("30" = "white", "37" = "grey9")

condition_outline_colors <- c("SD" = "#F8766D", "LN" = "#00BA38", "LS" = "#619CFF")

override_fill_for_legend <- unname(c( # Use unname() here
  temp_fill_colors["30"], temp_fill_colors["37"], # SD row
   temp_fill_colors["30"], temp_fill_colors["37"], # LN row
   temp_fill_colors["30"], temp_fill_colors["37"]  # LS row
))

override_color_for_legend <- unname(c( # Use unname() here
  condition_outline_colors["SD"],condition_outline_colors["SD"], # SD row
   condition_outline_colors["LN"],condition_outline_colors["LN"], # LN row
  condition_outline_colors["LS"],condition_outline_colors["LS"]  # LS row
))

```

```{r graphing multi filtered}
df_bw_sep %>% 
  ggplot(aes(x=conditions, y=abs))+
  geom_boxplot(aes(fill = conditions, color = Condition), outlier.shape = 15, outlier.color = "#F8766D", outlier.stroke =1)+
  geom_text(data = df_bw_outlier,
            aes(x = conditions, y = abs, label = label_text),
            color = "black",
            vjust = -.5,
             size = 3)+
  scale_fill_manual(
    name = "Condition / Temp",
    values = c(
      "SD30" = "grey100",
      "SD37" = "grey50",
      "LN37" = "grey50",
      "LN30" = "grey100",
      "LS30" = "grey100", 
      "LS37" = "grey50"
    ),
    guide = guide_legend(
      nrow = 3,
      ncol = 2,
      byrow = TRUE,
      override.aes = list(color = c( "#F8766D", "#F8766D", "#00BA38", "#00BA38", "#619CFF", "#619CFF"),  size = 3, alpha = 1)))+
  scale_color_manual(
    values = c("SD" = "#F8766D", "LN" = "#00BA38", "LS" = "#619cff"),
    guide = "none"
  )+
  theme_bw() +
  theme(
    legend.position = "right",
    legend.box = "horizontal",
    strip.text = element_blank(),
    legend.title = element_text(face = "bold"),
    plot.margin = margin(10, 10, 10, 10),
    panel.spacing = unit(0.5, "cm")
  ) 
```

#### `geom_boxplot` Line Types {#geombox-line-types}

You can also change the type of lines on the box plots. Though this does not have the same type of effect as `color` or `fill` where it allows for distinguishing of data colors. This change would be purely aesthetic. To change the type of lines, use the `linetype` argument. There are three main types that the lines can be: - "solid" - "dashed" - "dotted"

```{r Line Styling}
df_bw %>% 
  ggplot(
    aes(x = conditions, y = abs)
  )+
  geom_boxplot(linetype = "dashed")
```

The next change, while still under the aesthetics category, is a valuable tool to display information from your graph.

#### `log10` Boxplots {#log10-boxplots}

When representing gene expression, you might want to represent the absorbance values in your data on a logarhythmic scale. To do so, follow the example provided:

```{r Log 10 Box}
df_bw %>% 
  ggplot(aes(x=conditions, y=abs))+
  geom_boxplot()+
  scale_y_continuous(trans = "log10")
```

### Box Plot Parameters

These changes will not be simply aesthetic based, but will change the actual box plots themselves and their shapes. The first of these is the `width` argument.

#### Boxplot Fences

An important change to the boxplots that we need to do is create our upper and lower fences of our data. To do so, we will need to write a function to create the fences.

[Jacob]{style="color: orange;"}: explain what fences are in boxplots, and explain why we use `1.5` in the script as a multiplier and what it means.

Below, I will provide a script example to do so:

```{r Flag Creation}
whisker_data <- df_bw %>%
  group_by(conditions) %>%
  mutate(
    Q1 = quantile(abs, 0.25, na.rm = TRUE),
    Q3 = quantile(abs, 0.75, na.rm = TRUE),
    IQR = Q3 - Q1,
    lower_fence_val = Q1 - 1.5 * IQR,
    upper_fence_val = Q3 + 1.5 * IQR
  ) %>%
  summarise(
    whisker_upper = max(abs[abs <= unique(upper_fence_val)], na.rm = TRUE),
    whisker_lower = min(abs[abs >= unique(lower_fence_val)], na.rm = TRUE),
    upper_fence = unique(upper_fence_val),
    lower_fence = unique(lower_fence_val)
  ) %>%
  ungroup()
whisker_data$conditions <- as.factor(whisker_data$conditions)

# Use the factor levels to get numeric positions
whisker_data$conditions <- as.numeric(whisker_data$conditions)

```

Now that we have created the flags, we can apply them to our box plots. To do so, we will use the `geom_segment` function. This will create the "fences" for our box and whisker plots. An example is provided:

```{r Boxplot w/Flags}
df_bw %>% 
  ggplot(aes(x=conditions, y=abs))+
  geom_boxplot()+
  geom_segment(data = whisker_data ,
               aes(x=as.numeric(conditions) + 0.25,
                   xend=as.numeric(conditions) - 0.25,
                   y=whisker_lower,
                   yend=whisker_lower),
               inherit.aes = FALSE)+
  geom_segment(data=whisker_data,
               aes(x=as.numeric(conditions) + 0.25,
                   xend=as.numeric(conditions) - 0.25,
                   y=whisker_upper,
                   yend=whisker_upper))
```

#### Width of Box Plots

To change the width of box plots, in cases where data might be closer together, you will need to use the `width` argument. To do so, follow the example below:

```{r Width}
df_bw %>% 
  ggplot(
    aes(x = conditions, y=abs)
  )+
  geom_boxplot(width = 0.5)+
  geom_jitter(width = 0.2)
```

When you decrease the width of the box and whisker plots, you might also want to decrease the widths of the fences for our box plots. To do this, we will change the aesthetic x conditions of the `geom_segment` argument to reduce the widths of the fences. Below, I will change the x values from +/- 0.25 to +/- 0.1:

```{r Fence Widths}
df_bw %>% 
  ggplot(aes(x=conditions, y=abs))+
  geom_boxplot(width = 0.5, outlier.shape = NA)+
  geom_segment(data=whisker_data,
               aes(x=as.numeric(conditions) - 0.1,
                   xend=as.numeric(conditions)+0.1,
                   y=whisker_lower,
                   yend=whisker_lower))+
  geom_segment(data=whisker_data,
               aes(x=as.numeric(conditions) - 0.1,
                   xend=as.numeric(conditions) + 0.1,
                   y=whisker_upper,
                   yend=whisker_upper))+
  geom_jitter(width = 0.2)
```

There might be times where you will not have the same amount of data for every box in the box and whisker plot. In that case, you would use the next section, **Variable Widths**.

##### Variable Widths

There is another way to create a differing width box and whisker plot in `ggplot`. You can use `varwidth` to make the box widths proportional to the square roots of the number of observations per group. To do this, follow the example below:

```{r VarWidth}
df_bw %>% 
  ggplot(
    aes(x=conditions,y=abs)
  )+
  geom_boxplot(varwidth = TRUE)
  
```

The previous data set contains a perfectly even amount of data for each condition. This does not allow variable widths to be shown. For illustration, we created a data set that contains a varying density of data. The boxplot that follows shows what this new data set looks like when variable widths is activated. The following boxplot has a new data set that does activate varaible widths:

```{r VarWidth Data, echo=FALSE}
conditions <- c(rep("SD30",24), rep("SD37", 20), rep("LS30", 16), rep("LS37", 12), rep("LN30", 8), rep("LN37", 4))
                
                
num_rows <- 24  
min_value <-  0.1  
max_value <- 0.5  

num_rows_2 <- 20
num_rows_3 <- 16  
num_rows_4 <- 12
num_rows_5 <- 8
num_rows_6 <- 4

variable_values_1 <- runif(num_rows, min = min_value, max = max_value)
variable_values_2 <- runif(num_rows_2, min = min_value, max = max_value)
variable_values_3 <- runif(num_rows_3, min = min_value, max = max_value)
variable_values_4 <- runif(num_rows_4, min = min_value, max = max_value)
variable_values_5 <- runif(num_rows_5, min = min_value, max = max_value)
variable_values_6 <- runif(num_rows_6, min = min_value, max = max_value)

abs <- c(variable_values_1, variable_values_2, variable_values_3, variable_values_4, variable_values_5, variable_values_6)
df_var <- data.frame(conditions, abs)

view(abs)
```

```{r VarWidth Shown}
df_var %>% 
  ggplot(aes(x=conditions, y=abs))+
  geom_boxplot(varwidth = TRUE)
```

#### Creating a Median Line

When using a variable width boxplot, there might be times where you want to represent the median of your data set on the boxplot graph. To do so, you first need to find the overall median of your data set. The following example shows how:

```{r Median finding}
overall_median <- median(df_bw$abs)
```

Afterwards, we need to add a line of `geom_hline` so that we can create a solid horizontal line across our boxplots. This will be representative of your median of your WHOLE data frame, so it might not intersect medians of your box plots.

```{r Median representation}
df_bw %>% 
  ggplot(
    aes(x=conditions,y=abs)
  )+
  geom_boxplot(varwidth = TRUE)+
  geom_hline(yintercept = overall_median, color = "red")
```

There are other types of lines that could be created as well. For instance, there could be a mean line, a Q3 of the data, and the median of the data shown. Below, I will show the functions required to create these lines.

```{r Line value finding}
overall_median <- median(df_bw$abs)
overall_mean <- mean(df_bw$abs)
Q3 <- as.numeric(quantile(df_bw$abs, 0.75))
```

Afterwards, we need to add a line of `geom_hline` so that we can create a solid horizontal line across our boxplots. This will be representative of your median of your WHOLE data frame, so it might not intersect medians of your box plots.

```{r Median graph}
df_bw %>% 
  ggplot(
    aes(x=conditions,y=abs)
  )+
  geom_boxplot(varwidth = TRUE)+
  geom_hline(yintercept = overall_median, color = "red")+
  geom_hline(yintercept = overall_mean, color = "blue", linetype = "dashed", linewidth = 1)+
  geom_hline(yintercept = Q3, color = "purple", linetype = "dotted", linewidth = 2)
```

#### Coordinate Flip

The final customization option discussed in this document is coordinate flipping. To do this, you would add a line at the end of your box plot code that flips the x and y-axis. Remember that the ordering will be reversed, as the data read from left-to-right will now be in an ascending order. If you wish to change this, you can reverse the order in your coding of your data frame. Below is an example of the coordinate flipped graph.

```{r Coordinate Flip}
df_bw %>% 
  ggplot(aes(x=conditions, y=abs))+ 
  geom_boxplot(outlier.shape = NA)+
  coord_flip()
```

The way to reorder this is to specify which order you want your data to be in. an example is shown below:

```{r Ordering}
ordered_df_box <- df_bw %>% 
  mutate(conditions = fct_relevel(conditions, "LS37", "LS30", "LN37", "LN30", "SD37", "SD30"))

ordered_df_box %>% 
  ggplot(aes(x=conditions, y=abs))+
  geom_boxplot(outlier.shape = NA)+
  coord_flip()
```
