---
title: "Hydra"
date: last-modified
date-format: "MMM D"
author: "Dylan M. Ayres, Jacob A. Serrano, Dr. Amy Wiles"
---

# Libraries

```{r Libraries,echo=TRUE, results = 'hide', warning = FALSE}
library(ggplot2)
library(tidyverse)
library(scales)
library(kableExtra)
library(patchwork)
library(tidyr)
library(boot)
library(scales)
```

# Combinations

Within R there is the possibility to create a graph consisting of two already existing graphs. To do this, we must first understand how to create the two graphs you desire which can be seen in previous chapters, such as [**Pheatmap**](Pheatmap.qmd), [**Violin**](Violin.qmd), etc. In this example, we use a density plot and a whisker plot that we created using the following code.

## Outlier

```{r date conversion, include=FALSE, eval=FALSE}
# 1. Get the path of the file currently open in RStudio
current_path <- rstudioapi::getActiveDocumentContext()$path

# 2. Extract the directory and base filename
dir_name <- dirname(current_path)
base_name <- "Hydra" # The static part of your name

# 3. Create the new name with today's date (e.g., "Jan_22")
new_name <- paste0(base_name, ".qmd")
new_path <- file.path(dir_name, new_name)

# 4. Rename the file on the disk
if (current_path != new_path) {
  file.rename(current_path, new_path)
  message("File renamed to: ", new_name)
}
```

```{r data creation, echo = FALSE, results = 'hide', warning = FALSE}
Verification_Data <- readxl::read_excel("C:/Users/dylan/Documents/R development/Heatmap/Copy of Verification Data.xlsx")
df_box <- Verification_Data[-c(1:288),]
delete_1 <- "_rep1$"
delete_2 <- "_rep2$"
delete_3 <- "_hr$"

df_box <- df_box %>% 
  separate(col = verification_position_hr, into = c("ver_pos", "_hr"), sep = "_") %>%
  select(-matches(delete_1), -matches(delete_2), -matches(delete_3))

df_box <- df_box %>%
  rename_with(~ str_remove_all(., "^norm_"), everything()) %>%
  rename_with(~ str_remove_all(., "_avg$"), everything())


df_bw <- df_box %>% 
  pivot_longer(cols = !c(1:2),
               names_to = "conditions",
               values_to = "abs")


df_bw_1 <- df_bw %>% 
  head(8)


```

```{r Outliers (fixed), echo = FALSE, results = 'hide', warning = FALSE}
filtered_data_df <- df_bw %>% 
  filter(conditions %in% c("SD37", "SD30"))
  

ordered_df <- filtered_data_df %>% 
  mutate(conditions = fct_relevel(conditions, "SD37", "SD30"))

data_fence <- ordered_df %>%
  filter(conditions %in% c("SD37", "SD30")) %>% 
  group_by(conditions) %>%
  mutate(
    Q1 = quantile(abs, 0.25),
    Q3 = quantile(abs, 0.75),
    IQR = Q3 - Q1,
    lower_fence = Q1 - 1.5 * IQR,
    upper_fence = Q3 + 1.5 * IQR,
    is_outlier = abs < lower_fence | abs > upper_fence
  ) %>%
    summarise(
    whisker_up = max(abs[abs <= unique(upper_fence)], na.rm = TRUE),
    whisker_lo = min(abs[abs >= unique(lower_fence)], na.rm = TRUE),
    upper_fen = unique(upper_fence),
    lower_fen = unique(lower_fence)
  ) %>%
  ungroup()
```

```{r outliers}
data_outlier <- df_bw %>%
  group_by(conditions) %>%
  mutate(
    Q1 = quantile(abs, 0.25),
    Q3 = quantile(abs, 0.75),
    IQR = Q3 - Q1,
    lower_fence = Q1 - 1.5 * IQR,
    upper_fence = Q3 + 1.5 * IQR,
    is_outlier = abs < lower_fence | abs > upper_fence
  ) %>%
  ungroup()

outlier_data <- data_outlier %>%
  filter(is_outlier) %>%
  mutate(outlier_label = paste0(ver_pos))

value_min <- min(filtered_data_df$abs)
value_max <- max(filtered_data_df$abs)
```

```{r Combinations}
density_A <- density(filtered_data_df$abs[filtered_data_df$conditions == "SD30"])
density_B <- density(filtered_data_df$abs[filtered_data_df$conditions == "SD37"])

# Extract the 'y' (density) values from the calculated densities
y_values_A <- density_A$y
y_values_B <- density_B$y

# Combine and find the overall min and max density
all_y_values <- c(y_values_A, y_values_B)
min_density <- min(all_y_values)
max_density <- max(all_y_values)

x_range <- max(filtered_data_df$abs) - min(filtered_data_df$abs)

y_range <- max_density - min_density

appropriate_ratio <- x_range/y_range

p_boxplot <- ggplot(ordered_df,
    aes(x=abs,y=conditions, color = conditions))+
  geom_boxplot(outlier.shape = 15, outlier.stroke = 1, outlier.colour = "black")+
  xlim(value_min, value_max)+
  geom_segment(data=data_fence,
               aes(x=whisker_up,
                   xend=whisker_up,
                   y=as.numeric(conditions) - 0.25,
                   yend=as.numeric(conditions) + 0.25))+
geom_segment(data=data_fence,
               aes(x=whisker_lo,
                   xend=whisker_lo,
                   y=as.numeric(conditions) - 0.25,
                   yend=as.numeric(conditions) + 0.25))+
geom_text(data = outlier_data,
            aes(x = abs, y = conditions, label = ver_pos),
            color = "black",
            vjust = -1,
            size = 3)+
scale_color_manual(breaks = c("SD30", "SD37"), values = c("SD30" = "#F8766D","SD37" ="#A3A500"))+
  scale_x_continuous(limits = c(0.1,0.4))
  
p_density <- ggplot(filtered_data_df,
    aes(x=abs, color=conditions)
  )+
  facet_wrap(~ conditions, ncol = 1)+
  xlim(value_min, value_max)+
  geom_density(aes(colors = conditions), show.legend = FALSE)+
  scale_color_manual(values = c("#F8766D", "#A3A500"))+
  scale_x_continuous(limits = c(0.1, 0.4))+
  theme(strip.text = element_blank())

p_boxplot + p_density +plot_layout(nrow = 1, ncol = 2)

(p_boxplot / p_density)

combined_plot <- (p_boxplot / p_density +
  plot_layout(widths = c(6, 0.1), heights = c(1, 3)))

combined_plot
```

```{r dens in box}
filtered_data_df$dens <- filtered_data_df %>% 
  mutate(dens = 6.5)

filtered_data_df %>% 
  ggplot(x = abs, color = conditions)+
  facet_wrap(~ conditions, ncol = 1)+
  xlim(value_min, value_max)+
   geom_boxplot(aes(x = abs, y = 6.25, color = conditions), width = 2, outlier.shape = 15, outlier.color = "black")+
  geom_density(aes(x = abs, color = conditions), show.legend = FALSE)+
  scale_x_continuous(limits = c(0.1, 0.4))+
  labs(y = "density")+
  geom_segment(data=data_fence,
               aes(x=whisker_up,
                   xend=whisker_up,
                   y=as.numeric(6.25) - .75,
                   yend=as.numeric(6.25) + .75),
             color = (values = c( "#A3A500", "#F8766D")))+
geom_segment(data=data_fence,
               aes(x=whisker_lo,
                   xend=whisker_lo,
                   y=as.numeric(6.25) - .75,
                   yend=as.numeric(6.25) + .75),
             color = (values = c("#A3A500", "#F8766D")))+
geom_text(data = outlier_data,
            aes(x = abs, y = 6.25, label = ver_pos),
            color = "black",
            vjust = -1,
            size = 3)+
  scale_color_manual(values = c("#F8766D", "#A3A500"))+
  theme(strip.text = element_blank())

```
