---
title: "Pivoting"
date: last-modified
date-format: "MMM D"
author: "Dylan M. Ayres, Jacob A. Serrano, Dr. Amy Wiles"

---

# File Download & Packages

Before we can begin to work with data, we must first import data we wish to use and install packages to run the code we create. To do this, follow the instructions on [**File Download Instruction**](File_download.qmd).

Ensure that you have the `tidyverse` installed in RStudio to enable the functions found here, follow the example below.

```{r Tidyverse, eval=FALSE}
 install.packages("tidyverse")
library(tidyverse)
```

```{r Tidyverse non-vis, warning=FALSE, results='hide', message=FALSE, include = FALSE, echo=FALSE}
# install.packages("tidyverse")
library(tidyverse)
```

To read in the data table you wish to work with, follow instructions in [**File Download Instruction**](File_download.qmd) section 2, [**Reading Data**](File_download.qmd#reading-data).

# Creating a Data Frame

Once you have your data table that you plan to use, you can skip to section 3, `Pivoting a Data Set`. However, for educational purposes, I will create a sample data set to be used for the remainder of this document.

```{r date conversion, include=FALSE, eval=FALSE}
# 1. Get the path of the file currently open in RStudio
current_path <- rstudioapi::getActiveDocumentContext()$path

# 2. Extract the directory and base filename
dir_name <- dirname(current_path)
base_name <- "Pivot" # The static part of your name

# 3. Create the new name with today's date (e.g., "Jan_22")
new_name <- paste0(base_name, ".qmd")
new_path <- file.path(dir_name, new_name)

# 4. Rename the file on the disk
if (current_path != new_path) {
  file.rename(current_path, new_path)
  message("File renamed to: ", new_name)
}
```

```{r Data Creation}
Gene_ID <- c("2401A01", "2401A03", "2401A04")
Condition_1 <- c(0.01, 0.05, 0.1)
Condition_2 <- c(0.5, 0.1, 1)

Data <- data.frame(Gene_ID, Condition_1, Condition_2)

print (Data)
```

# Pivoting a Data Set

Now, we can start to manipulate our data set. In some instances, a package may require a data frame to be in a vertical or horizontal orientation. Generally, NCBI data will be formatted in a matrix which will be incompatible with `ggplot`. We can fix this using a method called "pivoting," where we can change our matrix into a format that works with `ggplot` aptly named a pivot table.

To start, use either the `pivot_longer`, or `pivot_wider` function to manipulate your data. Pay attention to which one is being used, because 'longer' essentially rotates wide horizontal data to a vertical format, whilst 'wider' makes vertical data become horizontal.

## Pivot Longer

Pivot longer is a function that selects columns to be turned into rows in your data matrix. When working with "pivot_longer," make sure you set it up correctly with the following: use `cols` to move any pre-existing columns in the matrix to a new column in the pivoted table, `names_to` in order to assign a name to a new column created, and the `values_to` function to create a name for the column that your table values will fill.

Small things to remember, when pivoting large data structures, use the `cols =` argument to specify the columns that you wish to be pivoted. If you wish for a column to not be pivoted, but all others to pivot, you can use the `!` in front of the name of the column you wish to remain unchanged.

In the below code, we will show you the original file, and the file output when pivoted. The resulting table might be labeled `tibble`, which is a type of modified data frame in R, it is still usable for all function types. Notice that the `Gene ID` column remains unchanged and that the names of the rows created by `pivot_longer` are now set to `Conditions` and their values to `Expression`.

```{r Pivot_Longer}
b <- Data

print(b)

f <- pivot_longer(b, 
                  cols = !"Gene_ID",
                  names_to = "Conditions",
                  values_to = "Expression")
print(f)
```

Object `f`, shown above is a long format table of a data set. This object will be referenced again in the section 3.B "Pivot Wider."

## Pivot Wider

If we have a data set that is longer than it is wide, then we could use `pivot_wider`. When using `pivot_wider`, its important to note that the syntax to use pivot wider is different from pivot longer.

Pivot_wider by default will use the columns specified when creating a longer data frame. When you choose a column to pivot, all names in the cells will be grouped by names, and turned into columns. To fill those new columns with data, simply use the `values_from` function to dictate the data that will fill the new cells of the created columns. We are working with the object `f` again, and here we will set up object `g` to be the width wise table.

```{r Pivot_wider}
print(f)

g <- pivot_wider(f,
                 names_from = "Conditions",
                 values_from = "Expression"
                 )

print(g)
```
