---
title: "Volcano Plots"
date: last-modified
date-fomrat: "MMM D"
author: "Dylan M. Ayres, Jacob A. Serrano, Dr. Amy Wiles"

---

# Chapter Overview

Volcano plots are ways of relating one variable's expression to another. In the case of bioinformatics, this tends to be the binary logarithm of the intensity ratio between two values on a graph. In our case, we will be plotting data from a yeast data set using conditions of limited nitrogen and limited sulfur. For our data set, there are also two calculations of the `p_value` or measure of significance of the data. As such, we will further divide our graphing into the the log10 of these `p_value` calculations. The first step to undergo before creating a volcano plot is to install the packages.

# Packages

For this type of graph, we will be using some of the same packages as previous chapters. All libraries are shown in the code window below

```{r packages, eval = FALSE}
install.packages("tidyverse")
library(tidyverse)
install.packages("ggplot2")
library(ggplot2)
install.packages("dplyr")
library(dplyr)
install.packages("ggrepel")
library(ggrepel)
install.packages("patchwork")
library(patchwork)
install.packages("gridExtra")
library(gridExtra)
install.packages("grid")
library(grid)
install.packages("scales")
library(scales)
```

```{r Tidyverse, include=FALSE, echo=FALSE, message=FALSE, results='hide', warning=FALSE, error=FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(ggrepel)
library(patchwork)
library(gridExtra)
library(grid)
library(scales)
```

```{r date conversion, include=FALSE, eval=FALSE}
# 1. Get the path of the file currently open in RStudio
current_path <- rstudioapi::getActiveDocumentContext()$path

# 2. Extract the directory and base filename
dir_name <- dirname(current_path)
base_name <- "Volcano" # The static part of your name

# 3. Create the new name with today's date (e.g., "Jan_22")
new_name <- paste0(base_name, ".qmd")
new_path <- file.path(dir_name, new_name)

# 4. Rename the file on the disk
if (current_path != new_path) {
  file.rename(current_path, new_path)
  message("File renamed to: ", new_name)
}
```

```{r ggplot2 Data, include=FALSE, echo=FALSE}
Volcano_data <- readxl::read_excel("C:/Users/dylan/Documents/R development/Volcano/Volcano_data.xlsx")

df_volcano <- Volcano_data %>% 
  filter(!(`duplicate notes` %in% c("superceded by 2.0 plate", "dup LS not sig; not hit", "dup LN not sig; not hit"))) %>% 
  select(1:3, contains("_p_values"), contains("M_values"))

filtered_tags <- c("	
collection note: incorrect; not a hit", "library fail; dup used", "superseded by 2.0 plate", "	
dup w diff outcomes; not hit")

df_volcano_filtered <- df_volcano %>% 
  filter(!(trimws(`duplicate notes`) %in% trimws(filtered_tags))) %>% 
  drop_na(ORF)
```

# Labeling Your Data

Before we begin creating volcano plots, it is important to ensure that we label our data to determine what values are significant and which values are not significant. To do this, we will use the `mutate()` function to add a new column to our dataset with the labels. The code window below shows how this is done.

```{r data labeling, eval=FALSE}
BSQN_LN_SD_volcano <- df_volcano_filtered %>% 
  mutate(diffexpressed = case_when(
    LN_SD_M_values > 1 & BSQN_p_values_LN < 0.05 ~ "Up",
    LN_SD_M_values < -1 & BSQN_p_values_LN < 0.05 ~ "Down",
    TRUE ~ "No"
  ))
```

Breaking down this code, we are taking a data set and adding a new column called `diffexpressed`. In that column, we are specifying that in the cases when the M values are greater than 1 and the p values are less than 0.05 (for significance) the value should be labeled "Up". This is to simply provide a label for upregulated data points in this condition, as compared to the control condition that are statistically significant. We also provide a specification for the opposite effect and label it "Down" to signify downregulation of a gene between the control condition and this condition. The last line states that if there is data that does not fit these two criteria lines, it should be labeled "No" for no difference. This will be useful for specifying points in the volcano plot to be colored differently to visually represent the upregulated genes.

```{r BSQN LN data, include=FALSE, echo=FALSE, message=FALSE, results='hide', warning=FALSE, error=FALSE}
BSQN_LN_SD_volcano <- df_volcano_filtered %>%
  mutate(diffexpressed = case_when(
    LN_SD_M_values > 1 & BSQN_p_values_LN < 0.05 ~ "Up",
    LN_SD_M_values < -1 & BSQN_p_values_LN < 0.05 ~ "Down",
    TRUE ~ "No"
  )) %>% 
  select(-`duplicate notes`, -BSQN_p_values_LS, -cellhts2_LS_p_values, -cellhts2_LN_p_values, -LS_SD_M_values)
```

# Creating a Volcano Plot

Now we will use the ggplot2 package functionality to create a volcano plot. To do this, we will use the `aes()` argument in the `ggplot()` function to orient our two sets of values. Following this, we will use the functions for `geom_hline()` and `geom_vline()` to create a visual guide to areas where data would be significant. The code example below shows all function used to create the violin plot

```{r Volcano BSQN LN, eval=FALSE}
BSQN_LN <-ggplot(BSQN_LN_SD_volcano, aes(x=LN_SD_M_values, y = -log10(BSQN_p_values_LN), color = diffexpressed))+
  geom_point(alpha = 0.8, size = 1.5)+
  geom_hline(yintercept = -log10(0.05), linetype= "dashed", color = "black", linewidth = 0.5)+
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", linewidth = 0.5)+
  scale_color_manual(values = c("Down" = "#7D3560", "No" = "black", "Up" = "#4E7705" ))+
  labs(x = "M_value",
       y = "-log10 p value",
       color = "Expression status")+
    theme_minimal()+
    theme(legend.position = "right")+
```

In the code window above, we specified the M values on the x axis, the -log10 of the p values on the y axis, and color to be related to the `diffexpressed` column values. After that, the `geom_hline()` function creates a dashed line as a distinction for significance at the -log 10 of 0.05. The `geom_vline()` functoin creates vertical dashed lines at -1 and 1 to represent where differences between the condition and the control condition in terms of expression. The `scale_color_manual()` function creates a custom scale that uses the expression status column created before to color up- and down-regulated expression of genes. The visual of this is shown below:

```{r image, echo=FALSE, results='hold'}
BSQN_LN <-ggplot(BSQN_LN_SD_volcano, aes(x=LN_SD_M_values, y = -log10(BSQN_p_values_LN), color = diffexpressed))+
  geom_point(alpha = 0.8, size = 1.5)+
  geom_hline(yintercept = -log10(0.05), linetype= "dashed", color = "black", linewidth = 0.5)+
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", linewidth = 0.5)+
  scale_color_manual(values = c("Down" = "#7D3560", "No" = "black", "Up" = "#4E7705" ))+
  labs(x = "M_value",
       y = "-log10 p value",
       color = "Expression status")+
    theme_minimal()+
    theme(legend.position = "right")

BSQN_LN
```

```{r Volcano BSQN LS, include = FALSE}
BSQN_LN <-ggplot(BSQN_LN_SD_volcano, aes(x=LN_SD_M_values, y = -log10(BSQN_p_values_LN), color = diffexpressed))+
  geom_point(alpha = 0.8, size = 1.5)+
  geom_hline(yintercept = -log10(0.05), linetype= "dashed", color = "black", linewidth = 0.5)+
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", linewidth = 0.5)+
  scale_color_manual(values = c("Down" = "#7D3560", "No" = "black", "Up" = "#4E7705" ))+
  labs(x = "",
       y = "-log10 p value",
       color = "Expression status")+
    theme_minimal()+
    theme(legend.position = "right")

BSQN_LS_SD_volcano <- df_volcano_filtered %>%
  mutate(diffexpressed = case_when(
    LS_SD_M_values > 1 & BSQN_p_values_LS < 0.05 ~ "Up",
    LS_SD_M_values < -1 & BSQN_p_values_LS < 0.05 ~ "Down",
    TRUE ~ "No"
  )) %>% 
  select(-`duplicate notes`, -BSQN_p_values_LN, -cellhts2_LS_p_values, -cellhts2_LN_p_values, -LN_SD_M_values)

BSQN_LS <- ggplot(BSQN_LS_SD_volcano, aes(x=LS_SD_M_values, y = -log10(BSQN_p_values_LS), color = diffexpressed))+
  geom_point(alpha = 0.8, size = 1.5)+
  geom_hline(yintercept = -log10(0.05), linetype= "dashed", color = "black", linewidth = 0.5)+
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", linewidth = 0.5)+
  scale_color_manual(values = c("Down" = "#7D3560", "No" = "black", "Up" = "#4E7705" ))+
  labs(x = "",
       y = "",
       color = "Expression status")+
    theme_minimal()+
    theme(legend.position = "right")

```

```{r Volcano cellhtts2 LN, include = FALSE}
Cellhts2_LS_SD_volcano <- df_volcano_filtered %>%
  mutate(diffexpressed = case_when(
    LS_SD_M_values > 1 & cellhts2_LS_p_values < 0.05 ~ "Up",
    LS_SD_M_values < -1 & cellhts2_LS_p_values < 0.05 ~ "Down",
    TRUE ~ "No"
  )) %>% 
  select(-`duplicate notes`, -BSQN_p_values_LN, -BSQN_p_values_LS, -cellhts2_LN_p_values, -LN_SD_M_values)

Cellhts2_LS <-ggplot(Cellhts2_LS_SD_volcano, aes(x=LS_SD_M_values, y = -log10(cellhts2_LS_p_values), color = diffexpressed))+
  geom_point(alpha = 0.8, size = 1.5)+
  geom_hline(yintercept = -log10(0.05), linetype= "dashed", color = "black", linewidth = 0.5)+
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", linewidth = 0.5)+
  scale_color_manual(values = c("Down" = "#7D3560", "No" = "black", "Up" = "#4E7705" ))+
  labs(x = "M value to Standard",
       y = "",
       color = "Expression status")+
    theme_minimal()+
    theme(legend.position = "right")

```

```{r Volcano Cellhts2 LN, include = FALSE}
Cellhts2_LN_SD_volcano <- df_volcano_filtered %>%
  mutate(diffexpressed = case_when(
    LN_SD_M_values > 1 & cellhts2_LN_p_values < 0.05 ~ "Up",
    LN_SD_M_values < -1 & cellhts2_LN_p_values < 0.05 ~ "Down",
    TRUE ~ "No"
  )) %>% 
  select(-`duplicate notes`, -BSQN_p_values_LN, -BSQN_p_values_LS, -cellhts2_LS_p_values, -LS_SD_M_values)

Cellhts2_LN <-ggplot(Cellhts2_LN_SD_volcano, aes(x=LN_SD_M_values, y = -log10(cellhts2_LN_p_values), color = diffexpressed))+
  geom_point(alpha = 0.8, size = 1.5)+
  geom_hline(yintercept = -log10(0.05), linetype= "dashed", color = "black", linewidth = 0.5)+
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", linewidth = 0.5)+
  scale_color_manual(values = c("Down" = "#7D3560", "No" = "black", "Up" = "#4E7705" ))+
  labs(x = "M value to Standard",
       y = "-log10 p value",
       color = "Expression status")+
    theme_minimal()+
    theme(legend.position = "right")
```

# Combined Graphs w/ Legends
  
  Sometimes, when there are multiple conditions, we might want to arrange graphs together to be able to examine multiple conditions. One way that we can do this is to use `grid.arrange()` to combine multiple objects of violin plots. In this case, we will be combining these graphs by conditions in the columns and by analysis type in the rows.An example of this is shown below:
  
```{r Combined graph w/ legend}
grid.arrange(
  BSQN_LN, BSQN_LS, Cellhts2_LN, Cellhts2_LS, 
  ncol = 2,
  top = textGrob("Limited Nitrogen                     Limited Sulfur", 
                 gp = gpar(fontsize = 14, fontface = "bold")),
  left = textGrob("Cellhts2                  BSQN", 
                  rot = 90, gp = gpar(fontsize = 14, fontface = "bold"))
)
```

  For the `grid.arrange()` function, the graphed objects are arranged from left to right in order of how we want the graphs to be shown. If we specify in `ncol =` that we want 2 columns, the first two objects will be arranged in left to right, then the last two objects will fill in a second row. The arguments afterward of `left` and `top` are labels added to the sides of the arranged graphs.\

  The arranged graphs all have individual legends and provide unnecessary clutter on the screen. We can use a different method graphing the combined plots to remove some of the unnecessary clutter.

# Combined Graphs w/ Unified Legend

## Row Labels  
  In this case, we are going to use aspects of the `patchwork` package to combine multiple plots and create a cleaner graph. Before we combine the graph, we first need to specify our row and column labels. To do so, we will create a new object under `ggplot` that contains an annotation. This will be used later to provide a title for our rows or columns. An example of a row label is shown below:
  
```{r Row Label, eval=FALSE}
row1_label <- ggplot() + annotate("text", x=0, y=0, label = "BSQN", fontface = 2, angle = 90, size = 5)+
  theme_void()
```

  After creating our row label, we need to create an object with a spacer so that we can display our row names and not interfere with our plots. To do this, we will create a new object called `left-col` to seperate row names from graphs. An example is shown below:
  
```{r Left_col, eval=FALSE}
left_col <- plot_spacer() / row1_label / row2_label + plot_layout(heights = c(0.1, 1, 1))
```

  This line of code specifies an empty cell followed by two row labels. Afterwards, the layouts of the spacer and labels were changed so that the empty space had a smaller height and the row labels displayed at their full length.\
  
## Column Labels  
  
  Now that we have created the row labels and spacers, we need to create the labels for columns. While creating the labels for our columns, we can also add the graphs we made previously to the new object. An example of this is shown below:
  
```{r col_nitrogen, eval=FALSE}  
col_nitrogen <- (ggplot() + annotate("text", x = 0, y = 0, label = "Limited Nitrogen", fontface = 2, size = 5)+ theme_void())/
    BSQN_LN/
    Cellhts2_LN+
    plot_layout(heights = c(0.1, 1, 1))
```

  This line of code creates a similar text annotation for the column name as the row names previously,and places the text over the `BSQN_LN` plot object.Then, the `Cellhts2_LN` object was placed under the annotation and other plot. Once again, the `plot_layout()` lines perform a similar action as in the row annotation.\ 

## Combining

  Finally, we need to combine our plots, and the annotation column together. We will use a similar strategy as the previous two objects. This time, we will combine the `left_col` object, `col_nitrogen` object, and `col_sulfur` obejct in a "side-by-side" manner. Then, we will add an argument caled `guides` to collect the legend. Finally, we will specify the total legend position.An example is shown below:

```{r final_plot, eval=FALSE}
final_plot <- (left_col | col_nitrogen | col_sulfur) + 
  plot_layout(widths = c(0.1, 1, 1), guides = 'collect') & 
  theme(legend.position = "top")
```

  A reason why we may choose to use this configuration for creating a combined graph, is that it provides more customization for both annotations to the graphs, and how we choose to combine our graphs. It also provides an opportunity to collect the legends of all graphs and relocate them as one combined legend to a different area of the graph window in R. An example of the result is shown below:

```{r COmbined Graph, echo=FALSE, results='hold'}
row1_label <- ggplot() + annotate("text", x=0, y=0, label="BSQN", fontface=2, angle=90, size=5) + theme_void()
row2_label <- ggplot() + annotate("text", x=0, y=0, label="Cellhts2", fontface=2, angle=90, size=5) + theme_void()

# 2. Build the Left "Label" Column
# spacer / r1 / r2
left_col <- plot_spacer() / row1_label / row2_label + plot_layout(heights = c(0.1, 1, 1))

# 3. Build the Nitrogen Column (Label + 2 plots)
col_nitrogen <- (ggplot() + annotate("text", x=0, y=0, label="Limited Nitrogen", fontface=2, size = 5) + theme_void()) / 
                BSQN_LN / 
                Cellhts2_LN + 
                plot_layout(heights = c(0.1, 1, 1))

# 4. Build the Sulfur Column (Label + 2 plots)
col_sulfur <- (ggplot() + annotate("text", x=0, y=0, label="Limited Sulfur", fontface=2, size = 5) + theme_void()) / 
              BSQN_LS / 
              Cellhts2_LS + 
              plot_layout(heights = c(0.1, 1, 1))

# 5. Stitch them all together side-by-side
final_plot <- (left_col | col_nitrogen | col_sulfur) + 
  plot_layout(widths = c(0.1, 1, 1), guides = 'collect') & 
  theme(legend.position = "top")

final_plot
```

# Labeling

  When examining statistically significant values in the volcano plots, outliers are clearly visible. To understand if these outlier data sets are consistent throughout testing, we could label these values. We will do this in a similar way to that of the labels for Scatter plots, and use `geom_text_repel()` function. Below is an example line of code that you could add to the rest of your volcano pliot code to create these outlier labels:
  
```{r outlier labels, eval=FALSE}
geom_text_repel(
  data = subset(BSQN_LN_SD__volcano, (LN_SD_M_values >10 | LN_SD_M_values <-10)&
                  (-log10(BSQN_p_values_LN) > -log10(0.05))),
  aes(label = ORF),
  size =3,
  fontace = "bold",
  max.overlaps = Inf,
  box.padding = 0.5,
  show.legend = FALSE
)
```

  Breaking down the code line above, we are plotting a peice of text, using the data table `BSQN_LN_SD_volcano`. From there, we are subdividing the data into M values of greater than 10 and less than -10, with a `-log10` p value greater than the `-log10` p value of 0.05. From there, we make the label for this text using the column `ORF`, in the data set. For the data set we are using here, the `ORF` column is the one that contains the gene names/IDs being tested. From there, we specify the font size and style. The last 3 lines create rules for how many labels can be on screen (`Inf` means infinite), spacing between labels to prevent overlap on screen (0.5 in this case), and should there be an extra legend shown (we turn this to `FALSE` to prevent visual clutter on the graph). A complete volcano plot with this extra line of code included is shown below:  

```{r lblBSQN_LN, echo=FALSE, results='hold'}
lbl_BSQN_LN <-ggplot(BSQN_LN_SD_volcano, aes(x=LN_SD_M_values, y = -log10(BSQN_p_values_LN), color = diffexpressed))+
  geom_point(alpha = 0.8, size = 1.5)+
  geom_text_repel(
    data = subset(BSQN_LN_SD_volcano, (LN_SD_M_values > 10 | LN_SD_M_values < -10)&
                    (-log10(BSQN_p_values_LN) > -log10(0.05))),
    aes(label = ORF),
    size = 3,
    fontface = "bold",
    max.overlaps = Inf,    # Ensures all requested labels show up
    box.padding = 0.5,     # Adds space around the text for better legibility
    show.legend = FALSE
  ) +
  geom_hline(yintercept = -log10(0.05), linetype= "dashed", color = "black", linewidth = 0.5)+
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", linewidth = 0.5)+
  scale_color_manual(values = c("Down" = "#7D3560", "No" = "black", "Up" = "#4E7705" ))+
  labs(x = "",
       y = "-log10 p value",
       color = "Expression status")+
    theme_minimal()+
    theme(legend.position = "right")

lbl_BSQN_LN
```

```{r Labeled points, include = FALSE}
lbl_BSQN_LN <-ggplot(BSQN_LN_SD_volcano, aes(x=LN_SD_M_values, y = -log10(BSQN_p_values_LN), color = diffexpressed))+
  geom_point(alpha = 0.8, size = 1.5)+
  geom_text_repel(
    data = subset(BSQN_LN_SD_volcano, (LN_SD_M_values > 10 | LN_SD_M_values < -10)&
                    (-log10(BSQN_p_values_LN) > -log10(0.05))),
    aes(label = ORF),
    size = 3,
    fontface = "bold",
    max.overlaps = Inf,    # Ensures all requested labels show up
    box.padding = 0.5,     # Adds space around the text for better legibility
    show.legend = FALSE
  ) +
  geom_hline(yintercept = -log10(0.05), linetype= "dashed", color = "black", linewidth = 0.5)+
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", linewidth = 0.5)+
  scale_color_manual(values = c("Down" = "#7D3560", "No" = "black", "Up" = "#4E7705" ))+
  labs(x = "",
       y = "-log10 p value",
       color = "Expression status")+
    theme_minimal()+
    theme(legend.position = "right")

lbl_BSQN_LS <- ggplot(BSQN_LS_SD_volcano, aes(x=LS_SD_M_values, y = -log10(BSQN_p_values_LS), color = diffexpressed))+
  geom_point(alpha = 0.8, size = 1.5)+
  geom_text_repel(
    data = subset(BSQN_LS_SD_volcano, (LS_SD_M_values > 5 | LS_SD_M_values < -5)&
                    (-log10(BSQN_p_values_LS) > -log10(0.05))),
    aes(label = ORF),
    size = 3,
    fontface = "bold",
    max.overlaps = Inf,    # Ensures all requested labels show up
    box.padding = 0.5,     # Adds space around the text for better legibility
    show.legend = FALSE
  ) +
  geom_hline(yintercept = -log10(0.05), linetype= "dashed", color = "black", size = 0.5)+
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", size = 0.5)+
  scale_color_manual(values = c("Down" = "#7D3560", "No" = "black", "Up" = "#4E7705" ))+
  labs(x = "",
       y = "",
       color = "Expression status")+
    theme_minimal()+
    theme(legend.position = "right")

lbl_Cellhts2_LS <-ggplot(Cellhts2_LS_SD_volcano, aes(x=LS_SD_M_values, y = -log10(cellhts2_LS_p_values), color = diffexpressed))+
  geom_point(alpha = 0.8, size = 1.5)+
  geom_text_repel(
    data = subset(Cellhts2_LS_SD_volcano, (LS_SD_M_values > 5 | LS_SD_M_values < -5)&
                    (-log10(cellhts2_LS_p_values) > -log10(0.05))),
    aes(label = ORF),
    size = 3,
    fontface = "bold",
    max.overlaps = Inf,    # Ensures all requested labels show up
    box.padding = 0.5,     # Adds space around the text for better legibility
    show.legend = FALSE
  ) +
  geom_hline(yintercept = -log10(0.05), linetype= "dashed", color = "black", size = 0.5)+
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", size = 0.5)+
  scale_color_manual(values = c("Down" = "#7D3560", "No" = "black", "Up" = "#4E7705" ))+
  labs(x = "M value to Standard",
       y = "",
       color = "Expression status")+
    theme_minimal()+
    theme(legend.position = "right")

lbl_Cellhts2_LN <- ggplot(Cellhts2_LN_SD_volcano, aes(x=LN_SD_M_values, y = -log10(cellhts2_LN_p_values), color = diffexpressed))+
  geom_point(alpha = 0.8, size = 1.5)+
  geom_text_repel(
    data = subset(Cellhts2_LN_SD_volcano, (LN_SD_M_values > 10 | LN_SD_M_values < -10)&
                    (-log10(cellhts2_LN_p_values) > -log10(0.05))),
    aes(label = ORF),
    size = 3,
    fontface = "bold",
    max.overlaps = Inf,    # Ensures all requested labels show up
    box.padding = 0.5,     # Adds space around the text for better legibility
    show.legend = FALSE
  ) +
  geom_hline(yintercept = -log10(0.05), linetype= "dashed", color = "black", size = 0.5)+
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", size = 0.5)+
  scale_color_manual(values = c("Down" = "#7D3560", "No" = "black", "Up" = "#4E7705" ))+
  labs(x = "M value to Standard",
       y = "-log10 p value",
       color = "Expression status")+
    theme_minimal()+
    theme(legend.position = "right")
```

# Combined w/ Labels

  Now that we created these legends in the volcano plots, we can combine the graphs together again and compare outlier values across conditions and/or normalization methods. For normalization methods, there may be less variance of outliers than across conditions, but it can be useful for determining erroneous data, or strong reacting genes to conditions. Below is a combined plot using these labeled volcano graphs:

```{r Combined Labeled, echo = FALSE, results = 'hold'}
row1_label <- ggplot() + annotate("text", x=0, y=0, label="BSQN", fontface=2, angle=90, size=5) + theme_void()
row2_label <- ggplot() + annotate("text", x=0, y=0, label="Cellhts2", fontface=2, angle=90, size=5) + theme_void()

left_col <- plot_spacer() / row1_label / row2_label + plot_layout(heights = c(0.1, 1, 1))

lbl_col_nitrogen <- (ggplot() + annotate("text", x=0, y=0, label="Limited Nitrogen", fontface=2, size = 5) + theme_void()) / 
                lbl_BSQN_LN / 
                lbl_Cellhts2_LN + 
                plot_layout(heights = c(0.1, 1, 1))

lbl_col_sulfur <- (ggplot() + annotate("text", x=0, y=0, label="Limited Sulfur", fontface=2, size = 5) + theme_void()) / 
              lbl_BSQN_LS / 
              lbl_Cellhts2_LS + 
              plot_layout(heights = c(0.1, 1, 1))

lbl_final_plot <- (left_col | lbl_col_nitrogen | lbl_col_sulfur) + 
  plot_layout(widths = c(0.1, 1, 1), guides = 'collect') & 
  theme(legend.position = "top")

lbl_final_plot
```


# Logarithmic Scale
  
  In the earlier chapter of Scatter plots, we discussed how to create a `log10` break for the x and y axes. The `log10` break for the y axis is still applicable in this situation, but not for the x axis. In the case of volcano plots, we will need to create a new type of `log2` scale for the x axis. Below is an example of how we can do that:
  
```{r log2 example, eval=FALSE}
log2_minor_breaks <- function(limits){
  max_log <- max(abs(limits), na.rm = TRUE)
  if (max_log == 0) return(0)
  powers <- 0:ceiling(max_log)
  minor_log_values <- unlist(lapply(powers, function(p) {
    low <- 2^p
    high <- 2^(p + 1)
    log2(seq(low, high, length.out = 5)) 
  }))
  all_breaks <- unique(c(-minor_log_values, 0, minor_log_values))
  all_breaks[all_breaks >= limits[1] & all_breaks <= limits[2]]
}
```

  Breaking down this section of code, we are creating a new function that gathers that gathers a minimum and maximum value of the table. While this function is running, it is also removing any `NA` data points. Then, it is creating a new object called `powers` where the value ranges from 0 to the maximum value of the data. Next, it creates a `minor_log` object to gather the lowest value on the `powers` list and the highest value on the `powers` list. Then, it subdivides that distance into 5 chuncks. The second to last line takes all the `minor_log` values and creates a range from positive `minor_log` to -`minor_log`. Finally, the last line tells the function to only apply to the start and end of the axis instead of the infinitie range.

```{r Log 10 (y), echo = FALSE, results = 'hold'}
log_minor_breaks <- function(limits) {
  start <- floor(min(limits))
  end <- ceiling(max(limits))
  breaks <- unlist(lapply(start:end, function(i) {
    log10(1:10) + i
  }))
  breaks[breaks >= limits[1] & breaks <= limits[2]]
}

log2_minor_breaks <- function(limits) {
  max_log <- max(abs(limits), na.rm = TRUE)
  if (max_log == 0) return(0)
  powers <- 0:ceiling(max_log)
  minor_log_values <- unlist(lapply(powers, function(p) {
    low <- 2^p
    high <- 2^(p + 1)
    log2(seq(low, high, length.out = 5)) 
  }))
  all_breaks <- unique(c(-minor_log_values, 0, minor_log_values))
  all_breaks[all_breaks >= limits[1] & all_breaks <= limits[2]]
}
```

  When inserting your new `log10` and `log2` functions into your volcano plot, make sure to specify which axis and that it is continuous. In this case, I will be showing an example of the `log2` function being applied to the x-axis. The following example is showing how to specify a range, major break values, and to create minor breaks using the `log2` function created earlier.
  
```{r log2 x, eval=FALSE}
scale_x_continuous(
    limits = c(-17, 17),
    breaks = seq(-16, 16, by = 2),
    minor_breaks = log2_minor_breaks)
```

```{r log Volcano, echo = FALSE, results = 'hold'}
log_BSQN_LN <-ggplot(BSQN_LN_SD_volcano, aes(x=LN_SD_M_values, y = -log10(BSQN_p_values_LN), color = diffexpressed))+
  geom_point(alpha = 0.8, size = 2)+
  geom_text_repel(
    data = subset(BSQN_LN_SD_volcano, (LN_SD_M_values > 10 | LN_SD_M_values < -10)&
                    (-log10(BSQN_p_values_LN) > -log10(0.05))),
    aes(label = ORF),
    size = 3,
    fontface = "bold",
    max.overlaps = Inf,    
    box.padding = 0.5,     
    show.legend = FALSE
  ) +
  geom_hline(yintercept = -log10(0.05), linetype= "dashed", color = "black", linewidth = 0.5)+
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", linewidth = 0.5)+
  scale_color_manual(values = c("Down" = "#7D3560", "No" = "black", "Up" = "#4E7705" ))+
  scale_y_continuous(
    limits = c(-1, 8), 
    breaks = seq(-1, 8, by = 1),
    minor_breaks = log_minor_breaks,
    labels = math_format(10^.x))+
  scale_x_continuous(
    limits = c(-17, 17),
    breaks = seq(-16, 16, by = 2),
    minor_breaks = log2_minor_breaks) +
  labs(x = NULL,
       y = "-log10 p value",
       color = "Expression status:")+
    theme_minimal()+
    theme(legend.position = "right",
          axis.title = element_text(size = 16, face = "bold"),
          axis.text = element_text(size = 12))

log_BSQN_LS <- ggplot(BSQN_LS_SD_volcano, aes(x=LS_SD_M_values, y = -log10(BSQN_p_values_LS), color = diffexpressed))+
  geom_point(alpha = 0.8, size = 2)+
  geom_text_repel(
    data = subset(BSQN_LS_SD_volcano, (LS_SD_M_values > 5 | LS_SD_M_values < -5)&
                    (-log10(BSQN_p_values_LS) > -log10(0.05))),
    aes(label = ORF),
    size = 3,
    fontface = "bold",
    max.overlaps = Inf,    # Ensures all requested labels show up
    box.padding = 0.5,     # Adds space around the text for better legibility
    show.legend = FALSE
  ) +
  geom_hline(yintercept = -log10(0.05), linetype= "dashed", color = "black", size = 0.5)+
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", size = 0.5)+
  scale_color_manual(values = c("Down" = "#7D3560", "No" = "black", "Up" = "#4E7705" ))+
    scale_y_continuous(
    limits = c(-1, 8), 
    breaks = seq(-1, 8, by = 1),
    minor_breaks = log_minor_breaks,
    labels = math_format(10^.x))+
  scale_x_continuous(
    limits = c(-17, 17),
    breaks = seq(-16, 16, by = 2),
    minor_breaks = log2_minor_breaks) +
  labs(x = "",
       y = "",
       color = "Expression status:")+
    theme_minimal()+
    theme(legend.position = "right",
          axis.title = element_text(size = 16, face = "bold"),
          axis.text = element_text(size = 12))

log_Cellhts2_LS <-ggplot(Cellhts2_LS_SD_volcano, aes(x=LS_SD_M_values, y = -log10(cellhts2_LS_p_values), color = diffexpressed))+
  geom_point(alpha = 0.8, size = 2)+
  geom_text_repel(
    data = subset(Cellhts2_LS_SD_volcano, (LS_SD_M_values > 5 | LS_SD_M_values < -5)&
                    (-log10(cellhts2_LS_p_values) > -log10(0.05))),
    aes(label = ORF),
    size = 3,
    fontface = "bold",
    max.overlaps = Inf,    # Ensures all requested labels show up
    box.padding = 0.5,     # Adds space around the text for better legibility
    show.legend = FALSE
  ) +
  geom_hline(yintercept = -log10(0.05), linetype= "dashed", color = "black", size = 0.5)+
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", size = 0.5)+
  scale_color_manual(values = c("Down" = "#7D3560", "No" = "black", "Up" = "#4E7705" ))+
    scale_y_continuous(
    limits = c(-1, 8), 
    breaks = seq(-1, 8, by = 1),
    minor_breaks = log_minor_breaks,
    labels = math_format(10^.x))+
  scale_x_continuous(
    limits = c(-17, 17),
    breaks = seq(-16, 16, by = 2),
    minor_breaks = log2_minor_breaks) +
  labs(x = "M value to Standard",
       y = "",
       color = "Expression status:")+
    theme_minimal()+
    theme(legend.position = "right",
          axis.title = element_text(size = 16, face = "bold"),
          axis.text = element_text(size = 12))

log_Cellhts2_LN <- ggplot(Cellhts2_LN_SD_volcano, aes(x=LN_SD_M_values, y = -log10(cellhts2_LN_p_values), color = diffexpressed))+
  geom_point(alpha = 0.8, size = 2)+
  geom_text_repel(
    data = subset(Cellhts2_LN_SD_volcano, (LN_SD_M_values > 10 | LN_SD_M_values < -10)&
                    (-log10(cellhts2_LN_p_values) > -log10(0.05))),
    aes(label = ORF),
    size = 3,
    fontface = "bold",
    max.overlaps = Inf,    # Ensures all requested labels show up
    box.padding = 0.5,     # Adds space around the text for better legibility
    show.legend = FALSE
  ) +
  geom_hline(yintercept = -log10(0.05), linetype= "dashed", color = "black", size = 0.5)+
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", size = 0.5)+
  scale_color_manual(values = c("Down" = "#7D3560", "No" = "black", "Up" = "#4E7705" ))+
    scale_y_continuous(
    limits = c(-1, 8), 
    breaks = seq(-1, 8, by = 1),
    minor_breaks = log_minor_breaks,
    labels = math_format(10^.x))+
  scale_x_continuous(
    limits = c(-17, 17),
    breaks = seq(-16, 16, by = 2),
    minor_breaks = log2_minor_breaks) +
  labs(x = "M value to Standard",
       y = "-log10 p value",
       color = "Expression status:")+
    theme_minimal()+
    theme(legend.position = "right",
          axis.title = element_text(size = 16, face = "bold"),
          axis.text = element_text(size = 12))
```

Now that we know how to add the `log10` and `log2` breaks to our graphs, we can create the combined graphs again and add our new minor breaks to them. An example is shown below:

```{r log 10 combined, echo=FALSE, results = 'hold', fig.width=12, fig.height=10, fig.fullwidth=TRUE}
row1_label <- ggplot() + annotate("text", x=0, y=0, label="BSQN", fontface=2, angle=90, size=5) + theme_void()
row2_label <- ggplot() + annotate("text", x=0, y=0, label="Cellhts2", fontface=2, angle=90, size=5) + theme_void()

left_col <- plot_spacer() / row1_label / row2_label + plot_layout(heights = c(0.1, 1, 1))

log_col_nitrogen <- (ggplot() + annotate("text", x=0, y=0, label="Limited Nitrogen", fontface=2, size = 5) + theme_void()) / 
                log_BSQN_LN / 
                log_Cellhts2_LN + 
                plot_layout(heights = c(0.1, 1, 1))

log_col_sulfur <- (ggplot() + annotate("text", x=0, y=0, label="Limited Sulfur", fontface=2, size = 5) + theme_void()) / 
              log_BSQN_LS / 
              log_Cellhts2_LS + 
              plot_layout(heights = c(0.1, 1, 1))

log_final_plot <- (left_col | log_col_nitrogen | log_col_sulfur) + 
  plot_layout(widths = c(0.1, 1, 1), guides = 'collect') &
  guides(color = guide_legend(override.aes = list(size = 5)))&
  theme(
    legend.position = "top",
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 14),
    legend.key.size = unit(1.5, "cm"))

log_final_plot
```


```{r test data, include=FALSE, echo=FALSE, message=FALSE, results='hide', warning=FALSE, error=FALSE}
test_volcano <- df_volcano_filtered %>%
  mutate(diffexpressed = case_when(
    LN_SD_M_values > 1 & BSQN_p_values_LN < 0.05 ~ "Up",
    LN_SD_M_values < -1 & BSQN_p_values_LN < 0.05 ~ "Down",
    TRUE ~ "No"
  )) %>% 
  rename(
    BSQN_LN_p = BSQN_p_values_LN,
    BSQN_LS_p = BSQN_p_values_LS,
    SD_LN_M = LN_SD_M_values,
    SD_LS_M = LS_SD_M_values,
    Cellhts2_LN_p = cellhts2_LN_p_values,
    Cellhts2_LS_p = cellhts2_LS_p_values
  ) %>% 
  pivot_longer(cols=matches("BSQN|cellhts2|SD"),
               names_to=c("Method", "Condition", "Metric"),
               names_sep = "_",
               values_to = "value") %>% 
  pivot_wider(names_from = Metric,
              values_from = value)

sd_reference <- test_volcano %>%
  filter(Method == "SD") %>%
  select(ORF, Condition, SD_M = M) # Rename M to SD_M for clarity

main_data <- test_volcano %>%
  filter(Method != "SD")

final_plot_data <- main_data %>%
  left_join(sd_reference, by = c("ORF", "Condition"))
  
method_names <- c("BSQN" = "BSQN Normalization", 
                  "Cellhts2" = "CellHTS2 Normalization")

condition_names <- c("LN" = "Limited Nitrogen Condition", 
                     "LS" = "Limited Sulfur Condition")

log_facet <- final_plot_data %>% 
  filter(Method != "SD" ) %>% 
  ggplot(aes(x=SD_M, y = -log10(p), color = diffexpressed))+
  geom_point(alpha = 0.8, size = 2)+
  geom_text_repel(
    data = subset(final_plot_data, (SD_M > 10 | SD_M < -10)&
                    (-log10(p) > -log10(0.05))),
    aes(label = ORF),
    size = 3,
    fontface = "bold",
    max.overlaps = Inf,    
    box.padding = 0.5,     
    show.legend = FALSE
  ) +
  geom_hline(yintercept = -log10(0.05), linetype= "dashed", color = "black", linewidth = 0.5)+
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", linewidth = 0.5)+
  scale_color_manual(values = c("Down" = "#7D3560", "No" = "black", "Up" = "#4E7705" ),
                     labels = c("Down" = "Downregulated", "No" = "Unchanged", "Up" = "Upregulated"))+
  scale_y_continuous(
    limits = c(-1, 8), 
    breaks = seq(-1, 8, by = 1),
    minor_breaks = log_minor_breaks,
    labels = math_format(10^.x))+
  scale_x_continuous(
    limits = c(-17, 17),
    breaks = seq(-16, 16, by = 2),
    minor_breaks = log2_minor_breaks) +
  labs(x = "M value to Standard",
       y = "-log10 p value",
       color = "Expression status:")+
    theme_bw()+
    theme(legend.position = "top",
          axis.title = element_text(size = 16, face = "bold"),
          axis.text = element_text(size = 12))+
  facet_grid(Method ~ Condition,
             labeller = labeller(
               Method = method_names,
               Condition = condition_names
             ))

log_facet
```
